Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-04 article-models-tt.tex
removed the rubbish rule, there may be more
small changes

\section{Models of type theory}
In this section we pursue the idea of what a general model of type theory is by
axiomatizing what you can do with them. We have the following ideas:
\begin{enumerate}
\item There are dependent models and sections thereof. Particular instances
of sections: extension, weakening and substitution (and something for identity
function?). And like the original
extension, weakening and substitution, they're going to be compatible with each
other. Therefore we state our theory of models as an extension of type theory
without constructors.
\item Type theory without constructors is a model of itself, the canonical model $\mctx$.
\item for any (family of) model(s) $A$, there is the family model $\mfam{A}$ which
is a family of models over $A$.
\item The terms of $\mctx$ should be precisely contexts. Terms of $\subst{\Gamma}{\mfam{\mctx}}$
should be families over $\Gamma$.
\item if a model and a family of models over it are given, there is an extended model.
If we extend $A$ by $\mfam{A}$, we get the Sierpinski model of $A$.
\item Likewise, models can be weakened and substituted and there are identity
functions.

\end{enumerate}
So the theory of models is going to be an extension of this theory with this
data.

\subsection{The basic ingredients of the theory of models}
We first introduce the basic ingredients of our abstract theory of models.
\begin{align*}
& \inference
  {}
  {\jctx{\mctx}}
  \tag{the canonical model}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jfam{{\Gamma}{A}}{\mfam{A}}}
  \tag{families}\\
& \inference
  {\jterm{\Gamma}{A}{x}}
  {\jterm{{\Gamma}{A}}{\mfam{A}}{\mfam{x}}}
  \tag{term action of $\mfam{}$}\\
& \inference
  {\jctx{\Gamma}}
  {\jfam{{\Gamma}{\mfam{\Gamma}}}{\mtm{\Gamma}}}
  \tag{terms}
\intertext{The empty context of a model $A$ is simply a term of $A$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{\Gamma}{A}{\tfemp{A}}}
  \tag{empty context}
\intertext{Extension is going to be a morphism from $\ctxext{A}{\mfam{A}}$ to
$A$ in context $\Gamma$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
  \tag{extension}
\intertext{The action of weakening on families
is a morphism from $\ctxwk{\mfam{A}}{\mfam{A}}$ to
$\mfam{\mfam{A}}$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{{{\Gamma}{A}}{\mfam{A}}}{\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}}
  \tag{weakening}
\intertext{The action of weakening on terms
is a morphism from $\ctxwk{\mfam{A}}{\mtm{A}}$ to $\mtm{\mfam{\mfam{A}}}$ over
$\tfwk{A}^0$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
    {{{\Gamma}{A}}{\mfam{A}}}
    {\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}
    {\ctxwk{\mfam{A}}{\mtm{A}}}
    {\mtm{\mfam{\mfam{A}}}}
    {\tfwk{A}^1}}
  \tag{weakening}
\intertext{The action of substitution by a term on families
is going to be a morphism from $\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}$ to 
$\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$
in context $\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}}
\intertext{Likewise, the action of substitution by a term on terms of those
families is going to be a morphism from $\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}$
to $\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}$ over $\tfsubst{A}^0$ in context
$\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}}
  {\tfsubst{A}^1}}
\intertext{The identity functions are coded by}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm
   {{{\Gamma}{A}}{\mfam{A}}}
   {\subst{{\idfunc[\mfam{A}]}{\tfwk{A}^0}}{\mtm{\mfam{\mfam{A}}}}}
   {\tfid{A}}}
\end{align*}
We will call the terms that we introduced here the model constructors.

@@

\subsubsection{Terms of families of extensions}
\begin{equation*}
\inference
  {\jterm{\Gamma}{A}{x}
   \jterm{\Gamma}{\subst{x}{P}}{u}}
  {\jtermeq
    {{{\Gamma}{A}}{P}}
    {\mfam{{A}{P}}}
    {\mfam{{x}{u}}}
    {{\ctxwk{P}{\mfam{x}}}{\mfam{u}}}}
\end{equation*}

\subsubsection{Terms of extensions}
\begin{equation*}
\inference
  {\jfam{{\Gamma}{A}}{P}}
  {\jfameq
    {{{{\Gamma}{A}}{P}}{\mfam{{A}{P}}}}
    {\mtm{{A}{P}}}
    {{\ctxwk{{P}{\mfam{A}}}{{P}{\mtm{A}}}}{\ctxwk{{P}{\mfam{A}}}{\mtm{P}}}}}
\end{equation*}

>>>>

\section{The type theory of models of type theory}
In this section we pursue the idea of what a general model of type theory is by
axiomatizing what you can do with them. We have the following ideas:
\begin{enumerate}
\item There are dependent models and sections thereof. Particular instances
of sections: extension, weakening and substitution (and something for identity
function?). And like the original
extension, weakening and substitution, they're going to be compatible with each
other. Therefore we state our theory of models as an extension of type theory
without constructors.
\item Type theory without constructors is a model of itself, the canonical model $\mctx$.
\item for any (family of) model(s) $A$, there is the family model $\mfam{A}$ which
is a family of models over $A$.
\item The terms of $\mctx$ should be precisely contexts. Terms of $\subst{\Gamma}{\mfam{\mctx}}$
should be families over $\Gamma$.
\item if a model and a family of models over it are given, there is an extended model.
If we extend $A$ by $\mfam{A}$, we get the Sierpinski model of $A$.
\item Likewise, models can be weakened and substituted and there are identity
functions.
\end{enumerate}
So the theory of models is going to be an extension of this theory with this
data. The theory we are about to describe can be seen as an elementary theory
of the category (with families (and terms)) of categories (with families (and terms)).

\subsection{The basic ingredients of the type theory of models}
We first introduce the basic ingredients of our abstract theory of models.
\begin{align*}
& \inference
  {}
  {\jctx{\mctx}}
  \tag{the canonical model}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jfam{{\Gamma}{A}}{\mfam{A}}}
  \tag{families}\\
& \inference
  {\jctx{\Gamma}}
  {\jfam{{\Gamma}{\mfam{\Gamma}}}{\mtm{\Gamma}}}
  \tag{terms}
\intertext{The following two rules essentially make a start with saying that every term is functorial
in the apropriate sense (there will be more rules contributing to this vies):}
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jterm{{{\Gamma}{A}}{\mfam{A}}}{\subst{f}{\mfam{P}}}{\mfam{f}}}\\
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jterm{{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}{\subst{\mfam{f}}{{f}{\mtm{P}}}}{\mtm{f}}}
\intertext{The empty context of a model $A$ is simply a term of $A$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{\Gamma}{A}{\tfemp{A}}}
  \tag{empty context}
\intertext{Extension is going to be a morphism from $\ctxext{A}{\mfam{A}}$ to
$A$ in context $\Gamma$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
  \tag{extension}
\intertext{The action of weakening on families
is a morphism from $\ctxwk{\mfam{A}}{\mfam{A}}$ to
$\mfam{\mfam{A}}$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{{{\Gamma}{A}}{\mfam{A}}}{\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}}
  \tag{weakening}
\intertext{The action of weakening on terms
is a morphism from $\ctxwk{\mfam{A}}{\mtm{A}}$ to $\mtm{\mfam{\mfam{A}}}$ over
$\tfwk{A}^0$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
    {{{\Gamma}{A}}{\mfam{A}}}
    {\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}
    {\ctxwk{\mfam{A}}{\mtm{A}}}
    {\mtm{\mfam{\mfam{A}}}}
    {\tfwk{A}^1}}
  \tag{weakening}
\intertext{The action of substitution by a term on families
is going to be a morphism from $\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}$ to 
$\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$
in context $\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}}
\intertext{Likewise, the action of substitution by a term on terms of those
families is going to be a morphism from $\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}$
to $\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}$ over $\tfsubst{A}^0$ in context
$\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}}
  {\tfsubst{A}^1}}
\intertext{The identity functions are coded by}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm
   {{{\Gamma}{A}}{\mfam{A}}}
   {\subst{{\idfunc[\mfam{A}]}{\tfwk{A}^0}}{\mtm{\mfam{\mfam{A}}}}}
   {\tfid{A}}}
\end{align*}
We will call the terms that we introduced here the model constructors.

@@

\subsubsection{Terms of families of extensions}

\subsubsection{Terms of extensions}

\subsubsection{Families over the empty context}
\begin{equation*}
\inference
  {\jfam{\Gamma}{A}}
  {\jfameq{\Gamma}{\subst{\tfemp{A}}{\mfam{A}}}{A}}
\end{equation*}

\subsubsection{Terms of the empty family}
\begin{equation*}
\inference
  {\jfam{\Gamma}{A}}
  {\jfameq{{\Gamma}{A}}{\subst{\tfemp{\mfam{A}}}{\mtm{A}}}{\emptyf[A]}}
\end{equation*}
